Graded Call-By-Push-Value

### 2020-07-08 popl21/popl21.tex
[ popl21 ] spell check run

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
 \begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010125.10010130</concept_id>
<concept_desc>Theory of computation~Type structures</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010138.10010142</concept_id>
<concept_desc>Theory of computation~Program verification</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010131.10010134</concept_id>
<concept_desc>Theory of computation~Operational semantics</concept_desc>
<concept_significance>100</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

@@

%\newcommand{}{\ensuremath{}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

\newcommand{\denp}{\den[+]}
\newcommand{\denn}{\den[-]}
\newcommand{\Denp}[2]{\denp{#1}_{#2}}
\newcommand{\Denn}[2]{\denn{#1}_{#2}}
\newcommand{\Den}[2]{\den{#1}_{#2}}
\newcommand{\denv}[2]{\dent{#1}_{#2}}

@@

The other introduction and elimination rules are just as in pure
CBPV, except that they propagate the effect classifier $e$ from
hypothesis to conclusion.  Note that in case distinction (\relim\GS)
and record construction (\rintro\Pi) all subterms $t_i$ are required
to produce effects classified by the same $e$.  However, in reality,
different branches of a case distinction, for instance, may produce
very different effects.  To end up with a unique classifier $e$, the
branches may have to be typed using effect subsumption (\rsub).  In
fact, the uses of \rsub can be confined to the hypotheses of \relim\GS
and \rintro\Pi, except for a final invokation of \rsub at the very end
of the typing derivation.  Alternatively, we could have introduced
effect algebras with suprema $\sup_{i:I} e_i$ instead of a preorder $e
\leq e'$.  However, suprema might not always exist; by using
subsumption \rsub, we delegate the problem of partiality to the
construction of a typing derivation.

@@

The coercion morphisms $\Denn N {e_1 \leq e_2}$ are obtained
analogously:
\[
\begin{array}{rcl}
  \Denn N {e_1 \leq e_2} & : & \Hom {\Denn N {e_1}} {\Denn N {e_2}} \\
  \Denn {\diamond P} {e_1 \leq e_2} & = & \tcast \\
  \Denn {P \To N}   {e_1 \leq e_2} & = &
    \Lambda(\Denn N {e_1 \leq e_2} \comp \teval) \\
  \Denn {\rect I N} {e_1 \leq e_2} & = &
    \langle \Denn N {e_1 \leq e_2} \comp \pi_i {} \rangle_{i:I} \\
\end{array}
\]
That these coercions satify the identity and composition---the functor
laws for $\denn N$---is easy to verify.

@@

If type constructor $\Box$ is interpreted as a monoidal comonad $\D$,
then all positive types $P$ can be intepreted as $\D$-coalgebras
$A = \den P$, \ie, implement a morphism $\expose[A] : A \to \D\,A$
satisfying the laws of a comonad coalgebra:
\[
\xymatrix@C=10ex{
A  \ar@{=}[dr]
  & \D\,A \ar[l]_{\textract} \ar[r]^{\tdisplay}
  & \D\,(\D\,A)
\\
  & A \ar[u]_{\texpose} \ar[r]^{\texpose}
  & \D\,A \ar[u]_{\tfmap\,\texpose}
}
\]
The implementation of $\expose[\den P]$ proceeds by induction on $P$.
\[
\begin{array}{lcl}
  \expose[\den{\Box N}] & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
  \expose[\den{\Box N}] & = & \display
\\[1ex]
  \expose[\den{\tupt I P}] & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
  \expose[\den{\tupt I P}] & = & \tzip \comp \tupt {i:I} \expose[\den{P_i}]
\\[1ex]
  \expose[\den{\sumt I P}] & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
  \expose[\den{\sumt I P}] & = & \left[\tfmap\,\iota_i \comp \expose[\den{P_i}]\right]_{i:I} \\
\end{array}
\]
Positive base types $o \in \PTy$ are required to be $\D$-coalgebras.
For instance, elements of a base type like
$\Float \in \PTy$ could represent streams of floating point numbers,
\eg, continuous measurements from a sensor.

@@

Graded comonads have been utilized to give semantics to
context-dependent computation \citep{orchard:icfp14}.  For grading,
loccit uses a \emph{resource algebra} in form of a pre-ordered
semiring
$(\R,{+},0,{\cdot},1,\leq)$.
% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
%
The semantics is \emph{resource aware} and thus not constructed in a
cartesian-closed category but in a symmetric monoidal closed category
$(\CC,\otimes,\I,\lolli)$.  We overload the operators on
exponentials, $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
$\otimes$ is a bifunctor and it is customary to overload $\otimes$ for
the functorial action
$f_1 \otimes f_2 : \Hom{A_1 \otimes A_2}{B_1 \otimes B_2}$ where
$f_i : \Hom{A_i}{B_i}$.  We shall also use $I$-ary products
$\tupt {i:I} {A_i}$ and their functorial action $\tupt {i:I} {f_i}$.
Further, the symmetric monoidal structure is usually witnessed by
isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
unit $\rho_A : \Hom {A \otimes I} A$, associativity
$\Ga_{A,B,C} : \Hom{A \otimes (B \otimes C)} {(A \otimes B) \otimes
  C}$
and swap $\sigma_{A,B} : \Hom {A \otimes B} {B \otimes A}$.  We shall
summarize combinations of these isomorphisms under the name
$\isot : \Hom {\tupt I A} {\tupt J B}$ when the multisets
$\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
deletion of $\I$s.

@@

We will also need to interpret context extension.  To this end, we
shall employ a structured product
$(C \boxtimes D)_{\gamma.\delta} = C_\gamma \otimes D_\delta$ where
$\gamma : \R^I$ and $\delta : \R^J$ and thus $C \boxtimes D$ is a
$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitely use the
isomorphism between $1$-structured $\D$-coalgebras $C : \R^1 \to \CC$
and $\R$-graded $\D$-coalgebras $A : \R \to \CC$ and define
\[
\den{\ext \Gamma x P}
  = \den \Gamma \boxtimes \denp P
  ,
\]
sweeping name issues under the carpet (to be handled by de Bruijn indices).

@@

Function types $\qfun r P N$ are now graded by a resource qualifier
$r$ that specifies how the function argument is to be used in the
function body.  (This type is often written $!_{r}P \lolli N$.)  In
quantitative typing with $\R = \Pot \bN$, qualifier $r$ gives the
possible usage quantities of the function argument, e.g. $\{1\}$ for
exactly one use (linear), $\{0,1\}$ for at most one use (affine),
$\{0\}$ for no use (constant), and $\bN$ for arbitrary use
(unrestricted).  A resource qualifier could also be a security level
(public or private), or a sensitivity level (a non-negative real)
\citep{reedPeirce:icfp10}.  If a lambda abstraction $\lam x t$ is
typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
the resource context (rule \rintro\lolli).  If a function
$t : \qfun r P N$ is applied (rule \relim\lolli), we need an
$r$-qualified argument $v : rP$.  We could have given a qualified
value typing judgement $\qValTy \gamma \Gamma v {rP}$ in the style of
\citet{mcBride:wadler60} that---in the quantitative
interpretation---provides $r$ copies of $v$ to be consumed by the
function.  Such a judgement could come with a scaling rule
\[
\ru{\qValTy \gamma \Gamma v {sP}
  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
\]
that allows to scale the production of $v$ by $r$ if the resources are
scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling instead
into the transition from values to computations.
Thus, $\relim\lolli$ receives an
argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
$\delta$ for the argument are scaled by $r$, summing the resource
requirements for the application to $\gamma + r\delta$.
%
Analogous scaling of the eliminatee is baked into the other value
eliminators (\relim\Box, \relim\GS, \relim\otimes), into the
\rlet-rules, and into \rintro\diamond.

@@

The equational theory of pure CBPV comprises $\beta$ and $\eta$ laws,
and \emph{sequencing laws} that syntactically express that negative
types are monad algebras \citep[Fig.~11]{levi:hosc06}.  The sequencing
laws contain a generalization of the associativity law for monads and
allow to permute a bind under a $\lambda$ or record construction.  In
the presence of $\eta$ for functions and records, they are
interderivable with the following permutation laws:
\[
\begin{array}{llcl}
\rpi\diamond
  & \bind {x_2} {(\bind {x_1} {t_1} {t_2})} {t_3}
  & = & \bind {x_1} {t_1} {(\bind {x_2} {t_2} {t_3})}
\\
\rpi\lolli
  & \app {(\bind x u t)} v
  & = & \bind x u {\app t v}
\\
\rpi\Pi
  & \proj i {(\bind x  u t)}
  & = & \bind x u {\proj i t}
\end{array}
\]
Note that permutations for let-bindings like
$\app{(\qlet r x v t)}w = \qlet r x v {\app t w}$ are instances of the
$\beta$-law $\qlet r x v t = \qsubst v r x t$.
%
Herein, \fbox{$\qsubst v r x t$} shall the denote the
(capture-avoiding) replacement $\subst v x t$ of a positive variable
$x$ annotated by $r$ by value $v$ in term $t$.
%
Permutations for value-eliminations like
$\app {(\qsplits r v {\bar x} t)} w = \qsplits r v {\bar x} {\app t
  w}$
are derivable (using the $\beta$-laws) from the $\eta$-laws, like
$\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
      x}} r z {t'}}$ for $\otimes$.

Our extension of CBPV features negative variables $y$ standing for
computations $u$, thus, we require a notion of computation
substitution $\qsubst u r t y$ that should have the following semantics.
\begin{align*}
&
  \ru{\qCompTy \delta \Gamma u N \qquad
      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r t y} {N'}}
\\ &
  \dent{\qsubst u r t y} = \dent t
    \comp (\tid[\den\Gamma \gamma]
             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
    \comp \duplicate[\den\Gamma]
\end{align*}

@@

Substitution for coeffect-graded type system has been worked
out in detail by \citet{atkeyWood:types19}.  It straightforwardly extends to
coeffect-graded CBPV.
%
A substitution $\sigma$ is a finite map from variable to terms,
mapping positive variables $x$ to values $v$ and negative variables
$y$ to computations $t$.  Substitution typing
\fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
$\Psi \in \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
vector $\Psi(z) \in \R^{\dom\Gamma}$ for each variable
$z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
\[
  \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qCompTy \gamma \Gamma t N
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\]
Capture-avoiding parallel substitution into values \fbox{$v \sigma$} and
computations \fbox{$t \sigma$} is defined as usual by recursion on the term.

The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
linear map $\R^{\dom \Delta} \to \R^{\dom \Gamma}$ and allows us to
state the substitution theorem.
\begin{theorem}[Substitution preserves typing]
  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
  \begin{enumerate}
  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\Psi\delta} \Gamma {v\sigma} P$.
  \item If $\qCompTy \delta \Delta t N$ then
    $\pCompTy {\Psi\delta} \Gamma {t\sigma} N$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Straightforward adaptation of \citet{atkeyWood:types19}.
\end{proof}
\begin{corollary}[Subject reduction]
  $\beta$-reduction of coeffect-graded CBPV preserves types.
\end{corollary}
In the pure (effect-free) version, every computation at value type
returns a value.
\begin{conjecture}[Canonicity of pure coeffect-graded CBPV]
  If $\vdash t : \diamond P$ then $t =_{\beta} \ret v$ for some $\vdash v : P$.
\end{conjecture}
This normalization result can be proved with standard techniques such
as reducibility candidates or normalization by evaluation (\cf
\citet{abelSattler:ppdp19}).

The denotation of a substitution $\SubstTy \Psi \Gamma \sigma \Delta$
is a natural transformation
$\denv \sigma \delta : \Hom {\Den \Gamma {\Psi\delta}} {\Den \Delta
  \delta}$
such that
$
\Den \Delta {\delta \leq \delta'}
\comp
\denv \sigma \delta
=
\denv \sigma {\delta'}
\comp
\Den \Gamma {\Psi\delta \leq \Psi\delta'}
$.
It is defined by recursion as follows:
\begin{align*}
& \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
& \dent \sempty
& : \Hom {\Den\Gamma 0} \I
& & = \tdrop[\den\Gamma]
\\[2ex]
& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
& \dent{\sext \sigma x v}
& : \Hom  {\Den\Gamma{\Psi\delta,r\gamma}} {\Den \Delta \delta \otimes \Denp P r}
& & = \denv \sigma \delta \otimes \denv v r
\\[2ex]
& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qCompTy \gamma \Gamma t N
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\end{align*}

>>>>

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
 \begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010125.10010130</concept_id>
<concept_desc>Theory of computation~Type structures</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010138.10010142</concept_id>
<concept_desc>Theory of computation~Program verification</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010131.10010134</concept_id>
<concept_desc>Theory of computation~Operational semantics</concept_desc>
<concept_significance>100</concept_significance>
</concept>
</Cs2012>
\end{CCSXML}

@@

%\newcommand{}{\ensuremath{}}
\newcommand{\bla}{\ensuremath{\mbox{$$}}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

\newcommand{\denp}{\den[+]}
\newcommand{\denn}{\den[-]}
\newcommand{\Denp}[2]{\denp{#1}_{#2}}
\newcommand{\Denn}[2]{\denn{#1}_{#2}}
\newcommand{\Den}[2]{\den{#1}_{#2}}
\newcommand{\Denpar}[2]{\Den{#1}{(#2)}}
\newcommand{\denv}[2]{\dent{#1}_{#2}}
\newcommand{\denvp}[2]{\dent{#1}_{(#2)}}

@@

The other introduction and elimination rules are just as in pure
CBPV, except that they propagate the effect classifier $e$ from
hypothesis to conclusion.  Note that in case distinction (\relim\GS)
and record construction (\rintro\Pi) all subterms $t_i$ are required
to produce effects classified by the same $e$.  However, in reality,
different branches of a case distinction, for instance, may produce
very different effects.  To end up with a unique classifier $e$, the
branches may have to be typed using effect subsumption (\rsub).  In
fact, the uses of \rsub can be confined to the hypotheses of \relim\GS
and \rintro\Pi, except for a final invocation of \rsub at the very end
of the typing derivation.  Alternatively, we could have introduced
effect algebras with suprema $\sup_{i:I} e_i$ instead of a preorder $e
\leq e'$.  However, suprema might not always exist; by using
subsumption \rsub, we delegate the problem of partiality to the
construction of a typing derivation.

@@

The coercion morphisms $\Denn N {e_1 \leq e_2}$ are obtained
analogously:
\[
\begin{array}{rcl}
  \Denn N {e_1 \leq e_2} & : & \Hom {\Denn N {e_1}} {\Denn N {e_2}} \\
  \Denn {\diamond P} {e_1 \leq e_2} & = & \tcast \\
  \Denn {P \To N}   {e_1 \leq e_2} & = &
    \Lambda(\Denn N {e_1 \leq e_2} \comp \teval) \\
  \Denn {\rect I N} {e_1 \leq e_2} & = &
    \langle \Denn N {e_1 \leq e_2} \comp \pi_i {} \rangle_{i:I} \\
\end{array}
\]
That these coercions satisfy the identity and composition---the functor
laws for $\denn N$---is easy to verify.

@@

If type constructor $\Box$ is interpreted as a monoidal comonad $\D$,
then all positive types $P$ can be interpreted as $\D$-coalgebras
$A = \den P$, \ie, implement a morphism $\expose[A] : A \to \D\,A$
satisfying the laws of a comonad coalgebra:
\[
\xymatrix@C=10ex{
A  \ar@{=}[dr]
  & \D\,A \ar[l]_{\textract} \ar[r]^{\tdisplay}
  & \D\,(\D\,A)
\\
  & A \ar[u]_{\texpose} \ar[r]^{\texpose}
  & \D\,A \ar[u]_{\tfmap\,\texpose}
}
\]
The implementation of $\expose[\den P]$ proceeds by induction on $P$.
\[
\begin{array}{lcl}
  \expose[\den{\Box N}] & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
  \expose[\den{\Box N}] & = & \display
\\[1ex]
  \expose[\den{\tupt I P}] & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
  \expose[\den{\tupt I P}] & = & \tzip \comp \tupt {i:I} \expose[\den{P_i}]
\\[1ex]
  \expose[\den{\sumt I P}] & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
  \expose[\den{\sumt I P}] & = & \left[\tfmap\,\iota_i \comp \expose[\den{P_i}]\right]_{i:I} \\
\end{array}
\]
Positive base types $o \in \PTy$ are required to be $\D$-coalgebras.
For instance, elements of a base type like
$\Float \in \PTy$ could represent streams of floating point numbers,
\eg, continuous measurements from a sensor.

@@

Graded comonads have been utilized to give semantics to
context-dependent computation \citep{orchard:icfp14}.  For grading,
\loccit uses a \emph{resource algebra} in form of a pre-ordered
semiring
$(\R,{+},0,{\cdot},1,\leq)$.
% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
%
The semantics is \emph{resource aware} and thus not constructed in a
cartesian-closed category but in a symmetric monoidal closed category
$(\CC,\otimes,\I,\lolli)$.  We overload the operators on
exponentials, $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
$\otimes$ is a bifunctor and it is customary to overload $\otimes$ for
the functorial action
$f_1 \otimes f_2 : \Hom{A_1 \otimes A_2}{B_1 \otimes B_2}$ where
$f_i : \Hom{A_i}{B_i}$.  We shall also use $I$-ary products
$\tupt {i:I} {A_i}$ and their functorial action $\tupt {i:I} {f_i}$.
Further, the symmetric monoidal structure is usually witnessed by
isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
unit $\rho_A : \Hom {A \otimes I} A$, associativity
$\Ga_{A,B,C} : \Hom{A \otimes (B \otimes C)} {(A \otimes B) \otimes
  C}$
and swap $\sigma_{A,B} : \Hom {A \otimes B} {B \otimes A}$.  We shall
summarize combinations of these isomorphisms under the name
$\isot : \Hom {\tupt I A} {\tupt J B}$ when the multisets
$\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
deletion of $\I$s.

@@

We will also need to interpret context extension.  To this end, we
shall employ a structured product
$(C \boxtimes D)_{\gamma.\delta} = C_\gamma \otimes D_\delta$ where
$\gamma : \R^I$ and $\delta : \R^J$ and thus $C \boxtimes D$ is a
$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitly use the
isomorphism between $1$-structured $\D$-coalgebras $C : \R^1 \to \CC$
and $\R$-graded $\D$-coalgebras $A : \R \to \CC$ and define
\[
\den{\ext \Gamma x P}
  = \den \Gamma \boxtimes \denp P
  ,
\]
sweeping name issues under the carpet (to be handled by de Bruijn indices).

@@

Function types $\qfun r P N$ are now graded by a resource qualifier
$r$ that specifies how the function argument is to be used in the
function body.  (This type is often written $!_{r}P \lolli N$.)  In
quantitative typing with $\R = \Pot \bN$, qualifier $r$ gives the
possible usage quantities of the function argument, e.g. $\{1\}$ for
exactly one use (linear), $\{0,1\}$ for at most one use (affine),
$\{0\}$ for no use (constant), and $\bN$ for arbitrary use
(unrestricted).  A resource qualifier could also be a security level
(public or private), or a sensitivity level (a non-negative real)
\citep{reedPierce:icfp10}.  If a lambda abstraction $\lam x t$ is
typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
the resource context (rule \rintro\lolli).  If a function
$t : \qfun r P N$ is applied (rule \relim\lolli), we need an
$r$-qualified argument $v : rP$.  We could have given a qualified
value typing judgement $\qValTy \gamma \Gamma v {rP}$ in the style of
\citet{mcBride:wadler60} that---in the quantitative
interpretation---provides $r$ copies of $v$ to be consumed by the
function.  Such a judgement could come with a scaling rule
\[
\ru{\qValTy \gamma \Gamma v {sP}
  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
\]
that allows to scale the production of $v$ by $r$ if the resources are
scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling instead
into the transition from values to computations.
Thus, $\relim\lolli$ receives an
argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
$\delta$ for the argument are scaled by $r$, summing the resource
requirements for the application to $\gamma + r\delta$.
%
Analogous scaling of the eliminatee is baked into the other value
eliminators (\relim\Box, \relim\GS, \relim\otimes), into the
\rlet-rules, and into \rintro\diamond.

@@

The equational theory of pure CBPV comprises $\beta$ and $\eta$ laws,
and \emph{sequencing laws} that syntactically express that negative
types are monad algebras \citep[Fig.~11]{levi:hosc06}.  The sequencing
laws contain a generalization of the associativity law for monads and
allow to permute a bind under a $\lambda$ or record construction.  In
the presence of $\eta$ for functions and records, they are
inter-derivable with the following permutation laws:
\[
\begin{array}{llcl}
\rpi\diamond
  & \bind {x_2} {(\bind {x_1} {t_1} {t_2})} {t_3}
  & = & \bind {x_1} {t_1} {(\bind {x_2} {t_2} {t_3})}
\\
\rpi\lolli
  & \app {(\bind x u t)} v
  & = & \bind x u {\app t v}
\\
\rpi\Pi
  & \proj i {(\bind x  u t)}
  & = & \bind x u {\proj i t}
\end{array}
\]
Note that permutations for let-bindings like
$\app{(\qlet r x v t)}w = \qlet r x v {\app t w}$ are instances of the
$\beta$-law $\qlet r x v t = \qsubst v r x t$.
%
Herein, \fbox{$\qsubst v r x t$} shall the denote the
(capture-avoiding) replacement $\subst v x t$ of a positive variable
$x$ annotated by $r$ by value $v$ in term $t$.
%
Permutations for value-eliminations like
$\app {(\qsplits r v {\bar x} t)} w = \qsplits r v {\bar x} {\app t
  w}$
are derivable (using the $\beta$-laws) from the $\eta$-laws, like
$\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
      x}} r z {t'}}$ for $\otimes$.

Our extension of CBPV features negative variables $y$ standing for
computations $u$, thus, we require a notion of computation
substitution $\qsubst u r t y$ that should have the following semantics
(see corollaries \ref{cor:tysinglesub} and \ref{cor:semsinglesub}).
\begin{align*}
&
  \ru{\qCompTy \delta \Gamma u N \qquad
      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}}
\\ &
  \dent{\qsubst u r y t} = \dent t
    \comp (\tid[\den\Gamma \gamma]
             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
    \comp \duplicate[\den\Gamma]
\end{align*}

@@

\subsection{Coeffect-graded CBPV: substitution and metatheory}

Substitution for coeffect-graded type system has been worked
out in detail by \citet{atkeyWood:types19,woodAtkey:linearity20}.
It straightforwardly extends to coeffect-graded CBPV.
%
A substitution $\sigma$ is a finite map from variable to terms,
mapping positive variables $x$ to values $v$ and negative variables
$y$ to computations $t$.  Substitution typing
\fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
$\Psi : \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
vector $\Psi(z) : \R^{\dom\Gamma}$ for each variable
$z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
\[
  \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qCompTy \gamma \Gamma t N
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\]
Capture-avoiding parallel substitution into values \fbox{$v \sigma$} and
computations \fbox{$t \sigma$} is defined as usual by recursion on the term.

The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
linear map
$(\delta : \R^{\dom \Delta}) \mapsto (\delta\Psi : \R^{\dom \Gamma})$
and allows us to state the substitution theorem.
\begin{theorem}[Substitution preserves typing]
  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
  \begin{enumerate}
  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\delta\Psi} \Gamma {v\sigma} P$.
  \item If $\qCompTy \delta \Delta t N$ then
    $\pCompTy {\delta\Psi} \Gamma {t\sigma} N$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Straightforward adaptation of \citet{atkeyWood:types19}.
\end{proof}
\begin{corollary}[Single substitution preserves typing]
  \label{cor:tysinglesub}
  \bla
  \begin{enumerate}
  \item If $\qValTy \delta \Gamma v P$ and
      $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst v r x t} {N}$.
  \item If $\qCompTy \delta \Gamma u N$ and
      $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$
      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}$.
  \end{enumerate}
\end{corollary}
%\begin{remark}
Analogous statements are available for the substitution into values
$\ValTy {\qext \gamma \Gamma r x P} w {P'}$ and
$\ValTy {\qext \gamma \Gamma r y N} w {P}$, however, they are not
invoked in the proof of subject reduction.
%\end{remark}
\begin{corollary}[Subject reduction]
  $\beta$-reduction of coeffect-graded CBPV preserves types.
\end{corollary}
In the pure (effect-free) version, every computation at value type
returns a value.
\begin{conjecture}[Canonicity of pure coeffect-graded CBPV]
  If $\vdash t : \diamond P$ then $t =_{\beta} \ret v$ for some $\vdash v : P$.
\end{conjecture}
This normalization result can be proved with standard techniques such
as reducibility candidates or normalization by evaluation (\cf
\citet{abelSattler:ppdp19}).
